---
title: "Hands-on Exercise 4: Spatial Weights and Applications"
date: "September 17, 2026"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
---

# 8.2 The Study Area and Data

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

# 8.3 Getting the Data Into R Environment

```{r}
# Import shapefile into r environment

hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")

# Import csv file into r environment

hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")

# Performing relational join

hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

# 8.4 Visualising Regional Development Indicator

```{r}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

# 8.5 Computing Contiguity Spatial Weights

```{r}
# Computing (QUEEN) contiguity based neighbours

wm_q <- poly2nb(hunan, queen=TRUE)
summary(wm_q)

wm_q[[1]]

hunan$County[1]

hunan$NAME_3[c(2,3,4,57,85)]

nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1

str(wm_q)

# Creating (ROOK) contiguity based neighbours

wm_r <- poly2nb(hunan, queen=FALSE)
summary(wm_r)

# Visualising contiguity weights

longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])

latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])

coords <- cbind(longitude, latitude)

head(coords)

plot(hunan$geometry, border="lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")

plot(hunan$geometry, border="lightgrey")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")

par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="Queen Contiguity")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
plot(hunan$geometry, border="lightgrey", main="Rook Contiguity")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

# 8.6 Computing distance based neighbours

```{r}
# Determine the cut-off distance

#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)

# Computing fixed distance weight matrix

wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

**Quiz Answer:** Each county has, on average, 3.68 neighbours within 62 km. It's simply 324 total links ÷ 88 counties.

```{r}
str(wm_d62)

table(hunan$County, card(wm_d62))

n_comp <- n.comp.nb(wm_d62)
n_comp$nc

table(n_comp$comp.id)

plot(hunan$geometry, border="lightgrey")
plot(wm_d62, coords, add=TRUE)
plot(k1, coords, add=TRUE, col="red", length=0.08)

par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="1st nearest neighbours")
plot(k1, coords, add=TRUE, col="red", length=0.08)
plot(hunan$geometry, border="lightgrey", main="Distance link")
plot(wm_d62, coords, add=TRUE, pch = 19, cex = 0.6)

# Computing adaptive distance weight matrix

knn6 <- knn2nb(knearneigh(coords, k=6))
knn6

str(knn6)

plot(hunan$geometry, border="lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

# 8.7 Weights based on IDW

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1/(x))
ids
```

# 8.8 Row-standardised Weights Matrix

```{r}
rswm_q <- nb2listw(wm_q, style="W", zero.policy = TRUE)
rswm_q

rswm_q$weights[10]

rswm_ids <- nb2listw(wm_q, glist=ids, style="B", zero.policy=TRUE)
rswm_ids

rswm_ids$weights[1]

summary(unlist(rswm_ids$weights))
```

# 8.9 Application of Spatial Weight Matrix

```{r}
# Spatial lag with row-standardized weights

GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag

nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1

lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan,lag.res)

head(hunan)

gdppc <- qtm(hunan, "GDPPC")
lag_gdppc <- qtm(hunan, "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp=1, ncol=2)

# Spatial lag as a sum of neighboring values

b_weights <- lapply(wm_q, function(x) 0*x + 1)
b_weights2 <- nb2listw(wm_q, 
                       glist = b_weights, 
                       style = "B")
b_weights2

lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")

lag_sum

hunan <- left_join(hunan, lag.res)

gdppc <- qtm(hunan, "GDPPC")
lag_sum_gdppc <- qtm(hunan, "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp=1, ncol=2)

# Spatial window average

wm_qs <- include.self(wm_q)

wm_qs[[1]]

wm_qs <- nb2listw(wm_qs)
wm_qs

lag_w_avg_gpdpc <- lag.listw(wm_qs, 
                             hunan$GDPPC)
lag_w_avg_gpdpc

lag.list.wm_qs <- list(hunan$NAME_3, lag.listw(wm_qs, hunan$GDPPC))
lag_wm_qs.res <- as.data.frame(lag.list.wm_qs)
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")

hunan <- left_join(hunan, lag_wm_qs.res)

hunan %>%
  select("County", 
         "lag GDPPC", 
         "lag_window_avg GDPPC") %>%
  kable()

w_avg_gdppc <- qtm(hunan, "lag_window_avg GDPPC")
tmap_arrange(lag_gdppc, w_avg_gdppc, asp=1, ncol=2)

# Spatial window sum

wm_qs <- include.self(wm_q)
wm_qs

b_weights <- lapply(wm_qs, function(x) 0*x + 1)
b_weights[1]

b_weights2 <- nb2listw(wm_qs, 
                       glist = b_weights, 
                       style = "B")
b_weights2

w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc

w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")

hunan <- left_join(hunan, w_sum_gdppc.res)

hunan %>%
  select("County", "lag_sum GDPPC", "w_sum GDPPC") %>%
  kable()

w_sum_gdppc <- qtm(hunan, "w_sum GDPPC")
tmap_arrange(lag_sum_gdppc, w_sum_gdppc, asp=1, ncol=2)
```
